A flexible integer linear programming formulation for scheduling clinician on-call service in hospitals

### 2019-08-29 paper/paper.discussion.tex
Work on discussion

Our formulation includes hard constraints to ensure the schedule satisfies hospital and logistics requirements. It also aims to satisfy the work preferences of clinicians in a clinical department by optimizing a multi-goal objective function. While one of our goals was to keep a minimal set of constraints to ensure that the ILP formulation is usable in practice, it leaves room to adapt it to various clinical departments at different hospitals. The flexibility of our ILP allows changing the number of services provided in a division, the length of a work block, clinicians' preference for work block to weekend adjacency as well as clinicians' requests for time off. \\

@@

We analyzed the performance of our formulation on simulated data meant to resemble real-life clinician departments at different hospitals. Our results show that increasing the number of requests per clinician does not affect the runtime of the algorithm, highlighting the flexibility of the tool in incorporating clinician preferences. Further, we saw that the algorithm can accommodate increasing the time-horizon up to four years with little impact on runtime, which can be applied to departments that generate schedules far in advance. The most notable impact on the runtime was noticed when trying to increase the number of services offered in a single department. While such cases are unlikely to be encountered in the real-world, as most departments tend to focus on a single service [ref \ref{???}], the runtime issues can be mitigated by relaxing certain constraints. \\

In future work, we can look first and foremost into improving the runtime of the tool by modifying constraints and objectives, so that it can be used in larger departments offering multiple services for their patients. On top of that, we can introduce new objectives that incorporate additional clinician preferences and allow for better distribution of work-load. However, since our ultimate goal is to maintain a ILP formulation, care must be taken not to overly complicate the model and risk running into additional time complexity issues.

>>>>

Our formulation includes hard constraints to ensure the schedule satisfies hospital and logistics requirements. It also aims to satisfy the work preferences of clinicians in a clinical department by optimizing a multi-goal objective function. While one of our goals was to keep a minimal set of constraints to ensure that the ILP formulation is usable in practice, it leaves room to adapt it to various clinical departments at different hospitals. The flexibility of our ILP allows changing the number of services provided in a division, the length of a work block, clinicians' preference for block to weekend adjacency as well as clinicians' requests for time off. \\

@@

We analyzed the performance of our formulation on simulated data meant to resemble real-life clinician departments at different hospitals. Our results show that increasing the number of requests per clinician does not affect the runtime of the algorithm, highlighting the flexibility of the tool in incorporating clinician preferences. Further, we saw that the algorithm can accommodate increasing the time-horizon up to four years with little impact on runtime, which can be applied to departments that generate schedules far in advance. The most notable impact on the runtime was noticed when trying to increase the number of services offered in a single department. While such cases are unlikely to be encountered in the real-world, as most departments tend to focus on a single service [ref \ref{???}], the runtime issues can be mitigated by relaxing certain constraints, which can then be readjusted manually from the generated schedule. Overall, it is very encouraging to observe that an ILP formulation can be applied to schedule clinicians in practice, by providing optimal schedules rather than having to utilize heuristics. \\

In future work, we can look first and foremost into improving the runtime of the tool by modifying constraints and objectives, so that it can be used in larger departments offering multiple services for their patients. On top of that, we can introduce new objectives that incorporate additional clinician preferences and allow for better distribution of work-load. However, since our ultimate goal is to maintain a ILP formulation, care must be taken not to overly complicate the model and avoid running into additional time complexity issues.
